---
title: "ggplot2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggplot2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Preview vignette with: devtools::build_rmd("vignettes/ggplot2.Rmd")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette shows you how to use palettes as colour and fill scales with ggplot2.

```{r setup}
library(palettes)
library(ggplot2)
```

## Colour scales

For discrete colours use `scale_colour_palette_d()`.

```{r}
ggplot(diamonds[sample(nrow(diamonds), 1000), ], aes(carat, price)) +
  geom_point(aes(colour = clarity)) +
  scale_colour_palette_d(met_palettes$VanGogh3)
``` 

Continuous or binned colours can be used with continuous data. For continuous colours use `scale_colour_palette_c()`. For binned colours use `scale_colour_palette_b()`.

```{r, out.width = "45%", fig.show = "hold"}
hwy_mpg <- ggplot(mpg, aes(displ, hwy, colour = hwy)) +
  geom_point()

hwy_mpg + scale_colour_palette_c(met_palettes$Greek)

hwy_mpg + scale_colour_palette_b(met_palettes$Greek)
```

## Fill scales

For discrete fills use `scale_fill_palette_d()`.

```{r}
ggplot(diamonds, aes(x = price, fill = cut)) +
  geom_histogram(position = "dodge", binwidth = 1000) +
  scale_fill_palette_d(pnw_palettes$Sunset)
```

Continuous or binned fills can be used with continuous data. For continuous fills use `scale_fill_palette_c()`. For binned fills use `scale_fill_palette_b()`.

```{r, out.width = "45%", fig.show = "hold"}
eruptions <- ggplot(faithfuld, aes(waiting, eruptions, fill = density)) +
  geom_tile()

eruptions + scale_fill_palette_c(viridis_palettes$mako)

eruptions + scale_fill_palette_b(viridis_palettes$mako)
```

<!-- ## Plot themes -->
