---
title: "Using palettes with gt"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using palettes with gt}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Preview vignette with: devtools::build_rmd("vignettes/gt.Rmd")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette shows you how to use palettes to colour data cells with **gt**.

```{r setup, message = FALSE, warning = FALSE}
library(palettes)
library(gt)
library(scales)
library(dplyr)
```

## Preparing the table

To demonstrate how to use palettes to colour data cells with gt, we will use `airquality` as an input data table with the following columns:

- Year, Month, Day: the numeric month and day of month for the record
- Temp: maximum daily air temperature in degrees Fahrenheit (°F)

Temperature is a good fit for colour---we can represent cold with blue and hot with red.

First we create the gt table we will add colour to. The table is slightly modified from a table in the [Introduction to Creating gt Tables](https://gt.rstudio.com/articles/intro-creating-gt-tables.html#the-column-labels) article. It looks like this.

```{r}
# Modify the `airquality` dataset by adding the year
# of the measurements (1973) and limiting to 10 rows
airquality_m <- 
  airquality %>%
  mutate(Year = 1973L) %>%
  slice(1:10) %>% 
  select(Year, Month, Day, Temp)
  
# Create a display table using the `airquality`
# dataset; arrange columns into groups
gt_tbl <- 
  gt(airquality_m) %>%
  tab_header(
    title = "New York Temperature Measurements",
    subtitle = "Daily measurements in New York City (May 1-10, 1973)"
  ) %>%
  tab_spanner(
    label = "Time",
    columns = c(Year, Month, Day)
  ) %>%
  tab_spanner(
    label = "Measurement",
    columns = c(Temp)
  ) %>%
  cols_label(
    Temp = html("Temp,<br>&deg;F")
  )

# Show the gt table
gt_tbl
```

## Adding colour

We can use the `Hiroshige` palette for blue and red gradients, reversing it so the colours are in the correct order for this example.

```{r}
colour_vector <- rev(met_palettes$Hiroshige)
colour_vector
```

At the moment, objects of class `palettes_palette` or `palettes_colour` cannot be used directly with the colour functions in gt. To work around this we cast the colour vector to a character vector.

```{r}
character_vector <- as.character(colour_vector)
character_vector
```

Now we can use `gt::data_color()` to colour the temperature cells. The `colors` argument accepts either a vector of colours to use for each distinct cell value or level or a colour mapping function (e.g., from the **scales** package).

Here we pass the character vector directly to the `colors` argument.

```{r}
gt_tbl %>% 
  data_color(
    columns = Temp,
    colors  = character_vector
  )
```

But this works equally well with the colour mapping functions from palettes.

```{r}
gt_tbl %>% 
  data_color(
    columns = Temp,
    colors  = pal_numeric(colour_vector, domain = NULL)
  )
```

The colour mapping functions from palettes are: `pal_numeric()`, `pal_bin()`, `pal_quantile()`, and `pal_factor()`. These functions are useful when you need finer control over colour evaluation with data.
